In the Fall of 2010, I was a doctoral student in Croatia and my research career was floundering. The dismal financial situation in my country had left me without funding and, although I was halfway resigned to a future of writing enterprise code, I decided to take one last gamble. I wrote an email to Mike Gleicher---whose work had greatly shaped my own research interests---asking if I could work for him as a visiting scholar. Mike introduced me to Bilge, then-junior HCI professor who had a thing for robots, gaze, and robots using gaze. While I knew little about gaze (and even less about robots), the project sounded interesting, and I certainly could not miss the opportunity to work with two scientists of such distinction. A few emails and a Skype chat later, I was buying a plane ticket to Wisconsin. And what was initially meant to be a brief scholarly visit would soon turn into an epic, six-year adventure in the land of ice and snow. We accomplished much during that time, yet none of it would have been been possible had my advisors not taken a gamble of their own that Fall. For that and much more, I thank them both. I also thank my other committee members---Eftychios Sifakis, Kevin Ponto, and Hrvoje Benko---for all the advice, suggestions, and critique they gave in the past few years. And finally, since this journey began in Zagreb, Croatia, I would be remiss not to thank Igor S. Pandzic, my advisor at the University of Zagreb, for his influence on my early growth as a researcher.

My path to a doctorate was exciting and often treacherous, and there are many who helped me along the way. The foremost among them is my coauthor, labmate, and dear friend Sean Andrist, who created the original version of the gaze model presented in Chapter~\ref{cha:GazeShiftModel} and whose advice and friendship got me through many hectic times in the last five years. Second in order but not in importance is my other coauthor and friend, Daniel Rakita (a.k.a., ``that guy''), who worked on the gaze authoring approach presented in Chapter~\ref{cha:GazeAuthoring} under my sometimes overbearing leadership. Besides my coauthors, many thanks go to others who contributed their time and skills to this research: Bobby Duncanson, Kevin Bennett, Elliot Hwang, Jessica Nelson, and other wonderful folks from Madison Media Institute for helping us obtain high-quality motion capture data; Arissa Sato for her annotation services; and Faye Golden for running my very last human subjects' study. I also extend my love and gratitude to all my labmates from the UW Visual Computing Lab and the HCI Lab, who provided me with drinking companionship as often as research feedback: Brandon Smith, Alper Sarikaya, Danielle and Dan Szafir, Michael Correll, Adrian Mayorga, Chris Bodden, Allison Saupp\'{e}, Margaret Pearce, Chien-Ming Huang, Irene Rae, Raj Setaluri, Zhi Tan, Eric Alexander, Nathan Mitchell, Liu Haixiang, Mike Doescher, Deidre Stuffer, Erin Winter, and others. You guys are the reason I will think back to this time with fondness and nostalgia. Last but not the least, I thank my family for all their love and support---especially my loving Mom. I know it was hard when your only son moved halfway across the world, but I hope I have made you proud with everything I achieved here. 